Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease

### 2023-05-12 doc/manuscript/translocation_pnas.tex
Make three panel model figure and update main text

% Please add a significance statement to explain the relevance of your work
\significancestatement{Diseases are becoming more common in wildlife and threaten many species
with extinction. TO BE WRITTEN}

@@

\begin{abstract}
Infectious diseases are emerging at an unprecedented rate, with
devastating consequences for global biodiversity. The recovery of
impacted taxa is hindered by the ongoing presence of disease, and few
broadly applicable conservation strategies exist to overcome this
obstacle. Rapid evolution in response to disease may produce
resistant/tolerant individuals capable of persisting despite the
presence of disease (``evolutionary rescue''), and although untested,
such hosts could be used to reestablish extirpated populations. We
evaluated this scenario using the once-common mountain yellow-legged
(MYL) frog, which is threatened with extinction by the emergence of a
lethal fungal pathogen. Although most MYL frog populations are
extirpated following disease outbreaks, some persist and eventually
begin to recover. We show that MYL frogs have undergone substantial
evolutionary change following disease outbreaks, and that these changes
are associated with increased resistance/tolerance to Bd infection in
naturally recovering populations. Large-scale reintroduction of frogs
from these populations resulted in the establishment of reproducing
populations at most recipient sites despite ongoing disease, and results
from viability modeling suggests that established populations have a low
probability of extinction over 50 years. Collectively, these results
provide one of the first examples of how evolution and the
reintroduction of resistant/tolerant individuals can allow the
landscape-scale recovery of disease-impacted taxa. This example has
potentially broad implications for the many taxa worldwide that are
threatened with extinction by emerging diseases.
\end{abstract}

@@

\dropcap{E}merging infectious diseases pose a severe threat to wildlife
populations \citep{daszak2000}, and have caused dramatic declines and
even extinctions in mammals, birds, amphibians, echinoderms, and other
taxa \citep{hewson2014, samuel2015, scheele2019, cunningham2021}.
Importantly, some populations previously decimated by disease have
subsequently recovered to varying degrees
\citep{newell2013, voyles2018, knapp2016} and might provide critical
insights for species conservation efforts. Unfortunately, given the
rarity of wildlife recoveries in the presence of disease, we know little
about the mechanisms driving population recovery
\citep{brannelly2021, russell2020} and how these mechanisms can be
leveraged to re-establish host populations across large spatial scales
\citep{mendelson2019}. At a time when novel infectious diseases are
emerging at an unprecedented rate \citep{daszak2000, fisher2012}, this
knowledge gap has far-reaching consequences for global biodiversity.

Host evolution following disease-induced declines is one mechanism by
which hosts can recover \citep{carlson2014, searle2020}. ``Evolutionary
rescue'' describes rapid evolutionary change that increases the
frequency of adaptive alleles and thereby restores positive population
growth and prevents extinction \citep{carlson2014}. Evolutionary changes
following disease-induced declines are known from a diversity of
wildlife populations
\citep{savage2016, epstein2016, gignoux-wolfsohn2021, holland2022}, and
often result in increased host resistance and/or tolerance. However,
linking these selective responses to population recoveries - as
necessary for evolutionary rescue - is difficult and remains rare. In
addition to promoting recovery of disease-impacted populations, the
resistant/tolerant individuals in recovering populations can, in theory,
be used to reestablish populations that were extirpated following
disease emergence \citep[e.g.,][]{mendelson2019}. To date, the use of
resistant/tolerant individuals for species recovery is rare \citep[but
see][]{joseph2018}, and studies that evaluate its potential utility are
critically needed.

There are four conditions that will often determine the feasibility of
using individuals from populations that are recovering following
disease-induced declines to reestablish extirpated populations. First,
populations that survive following disease emergence evolve increased
resistance/tolerance. Second, increased resistance/tolerance contributes
to higher individual fitness in the presence of disease and allows
positive population growth. Third, the fitness benefits gained through
selection are maintained beyond the natal environment and across
generations, allowing for the establishment of reintroduced populations.
And fourth, reintroduced populations show long-term growth and
persistence. To date, few of these conditions are documented, and
reintroducing hosts from recovering populations is largely untested
\citep{scheele2021, mendelson2019}.

The mountain yellow-legged (MYL) frog, composed of the sister species
\emph{Rana muscosa} and \emph{Rana sierrae} \citep{vredenburg2007}, is
emblematic of the unprecedented global declines of amphibians caused by
the recently-emerged amphibian chytrid fungus
\citep[\emph{Batrachochytrium dendrobatidis,}][]{scheele2019}. Once the
most common amphibian in the high elevation portion of the California's
Sierra Nevada mountains \citep[USA,][]{grinnell1924}, this frog has
disappeared from more than 90\% of its historical range during the past
century \citep{vredenburg2007}. Due to the severity of its decline and
the increasing probability of extinction, both species are now listed as
``endangered'' under the U.S. Endangered Species Act. In the Sierra
Nevada, this decline was initiated by the introduction of non-native
trout into the extensive historically-fishless region
\citep{bradford1989, knapp2000}, and the arrival of \emph{B.
dendrobatidis} (Bd) in the 1960s and its subsequent spread
\citep{vredenburg2019} caused additional widespread population
extirpations \citep{vredenburg2010, rachowicz2006}. These Bd-caused
declines are fundamentally different from the fish-caused declines
because removal of fish populations is feasible and results in the rapid
recovery of frogs \citep{knapp2007, vredenburg2004}. In contrast, Bd
appears to persist in habitats even in the absence of amphibian hosts
\citep{walker2007}, and therefore represents a long-term alteration of
invaded ecosystems that amphibians will need to overcome to reestablish
populations.

Despite the catastrophic impact of Bd on MYL frogs in which most
Bd-naive populations are extirpated following Bd arrival
\citep{vredenburg2010}, some populations have persisted after epizootics
\citep{briggs2010} and are now recovering \citep{knapp2016}. Frogs in
these recovering populations show reduced susceptibility to Bd
infection, consistent with selection for more resistant/tolerant
genotypes \citep{knapp2016}. Furthermore, exploratory efforts to
reintroduce frogs collected from recovering populations to sites from
which frogs were previously extirpated indicate that population
establishment is possible despite ongoing Bd infection
\citep{joseph2018}.

In the current study, we combine results from genomic analyses, a
15-year landscape-scale frog reintroduction effort, and detailed
population models to provide an unprecedented example of how evolution
in response to disease allows the successful reestablishment of
extirpated MYL frog populations. Specifically, we used exome capture
methods to show that MYL frogs have undergone substantial evolutionary
change associated with Bd exposure and that may provide a genetic basis
for the observed increased resistance/tolerance to Bd infection observed
in naturally recovering populations \citep{knapp2016}. We then describe
the reintroduction of these resistant/tolerant frogs to 24 sites and the
dynamics and drivers of population establishment using results from
long-term capture-mark-recapture (CMR) surveys. These data are also used
to parameterize a population model that extends our inferences of
population establishment well beyond the multi-year extent of the CMR
data. Collectively, these results provide one of the first examples of
how evolution and the reintroduction of resistant/tolerant animals can
allow the large-scale recovery of disease-impacted species. This example
exhibits several of the key characteristics of evolutionary rescue, and
has potentially broad implications for the many taxa worldwide that are
threatened with extinction by emerging diseases.

@@

Even when incorporating demographic and environmental stochasticity in
the probability of juvenile recruitment to the adult class \(\omega\)
(the transition that we expect to be the most subject to environmental
stochasticity in the presence of Bd), populations with high adult
survival are likely to persist over a 50 year time horizon. Our model
predicted that, following a single introduction of 40 adult individuals
into a population and using the baseline value for \(\omega\) of 0.3,
the six populations with the highest adult survival probabilities had
50-year extinction probabilities of 0 to 0.32;
Figure~\ref{fig-viability} B). This indicates strong potential for
long-term persistence in the presence of Bd and environmental
variability in recruitment (Figure~\ref{fig-viability} B). When we
decreased \(\omega\) from 0.3 to a conservative estimate of 0.15, 50
year extinction probability remained less than 0.5 for three populations
and was 0.79 to 0.94 for three additional populations
(Figure~\ref{fig-viability} B). In contrast, for the six populations
where \(\sigma_{A_R} < 0.5\), extinction probability over 50 years was
always predicted to be 1. To test the validity of our model predictions,
we demonstrated that our stochastic model could describe the general
recovery trajectory of our translocated population with the longest
survey history (Figure~\ref{fig-viability-supp} B; population 70550,
surveyed for 16 years).

@@

Disease-induced population declines are impacting global biodiversity
\citep{daszak2000}, and strategies to recover affected species are
lacking. Here, we used populations of resistant/tolerant individuals to
reestablish extirpated populations, despite ongoing disease. Although
previously proposed \citep{mendelson2019}, this strategy has rarely been
successfully implemented (see \citep{joseph2018} for a small-scale
example). In particular, although numerous studies have documented
signatures of selection in hosts following disease-induced declines
\citep[e.g.,][]{epstein2016, gignoux-wolfsohn2021}, we are unaware of
any that used individuals bearing the resulting adaptive alleles to
reestablish extirpated populations. The results of our study provide
evidence that (i) MYL frogs have experienced strong selection following
exposure to a recently-emerged lethal fungal disease (Bd) and that some
of the genomic regions under selection are associated with genes related
to disease resistance/tolerance, and (ii) reintroduction of individuals
from these resistant/tolerant populations allows the successful
reestablishment of viable frog populations despite ongoing Bd infection.
Although these results are generally consistent with expectations under
evolutionary rescue \citep{carlson2014}, obtaining strong evidence in
wild animal populations for each of the associated causal links (i.e.,
selection for adaptive alleles \(\rightarrow\) increase in host fitness
in the presence of the pathogen \(\rightarrow\) positive population
growth and recovery) is difficult and perhaps impossible. Nonetheless,
in the following narrative, we use evolutionary rescue as an
informative, albeit tentative framework.

@@

Results from our genomic analyses strongly suggest that natural
selection for adaptive alleles and resulting increased
resistance/tolerance is at least partially responsible for the reduced
susceptibility of frogs in recovering populations. We identified
specific alleles and genomic regions showing signatures of selection
between adjacent Bd-naive and recovering MYL frog populations,
consistent with strong selection following Bd exposure. Importantly,
some regions under selection are associated with cellular and
immunological mechanisms known to contribute to disease resistance,
including in amphibians \citep{zamudio2020}. For example, the MHC plays
an important role in immunity. In our study, we identified a region that
shows evidence of selection in recovering populations and contains eight
genes associated with either the MHC Class I or Class III regions. These
results corroborate numerous previous studies linking MHC genes to
amphibian resistance/tolerance against Bd
\citep[e.g.,][]{savage2011, bataille2015}. Similarly, the region with
the strongest indication of directional selection (as measured by
\(pi_{diff}\)) contains the gene GVINP1. Several previous studies have
found this gene to be differentially expressed during Bd infection
\citep[e.g.,][]{grogan2018, ellison2014}. This gene was also strongly
linked to disease in salmon, explaining a remarkable 20\% of the
resistance phenotype \citep{robledo2020, robledo2018}. Finally, the RIN3
gene showed a consistent pattern of allele frequency differences across
all nine of the frog populations sampled in this study, indicating
consistent selection in populations distributed across a wide geographic
area. In the SI, we provide additional details on these and other genes
identified in our study as under selection .

@@

Despite the resistance/tolerance of adult MYL frogs in recovering
populations, individual and population-level impacts of Bd are still
evident. In an earlier study of 2 of our 12 translocated populations, Bd
infection and load had detectable effects on the survival of adults and
may have influenced population establishment \citep[sites referred to as
``Alpine'' and ``Subalpine'' are referenced as ``70550'' and ``70505''
in the current study]{joseph2018}. Applying similar analyses to all 12
of our translocated populations would likely provide a broader
perspective of the ongoing effect of Bd. In addition to these important
but relatively subtle effects on adults, MYL frogs immediately following
metamorphosis (``metamorphs'') are highly susceptible to Bd infection
\citep{ellison2018} and experience high mortality. This high
susceptibility of metamorphs is documented for numerous species of
anurans, and may result from the poorly developed immune system
characteristic of this early life stage \citep{humphries2022}. In
naturally recovering and translocated MYL frog populations, we suggest
that the high mortality of metamorphs is an important limitation on
subsequent recruitment of new adults. Therefore, although adult MYL
frogs appear relatively resistant/tolerant, Bd infection continues to
have important limiting effects on recovering populations \citep[see
also][]{hollanders2022}.

The recent emergence of Bd worldwide has contributed to the decline of
hundreds of amphibian species, some of which are now extinct in the wild
\citep{scheele2019}. This unprecedented impact on global amphibian
biodiversity is compounded by the lack of any effective and broadly
applicable strategies to reverse these impacts
\citep{garner2016, knapp2022}. Importantly, in addition to the natural
recovery documented for MYL frogs \citep{knapp2016}, other species are
also showing evidence of post-epizootic recovery in the presence of Bd
\citep{scheele2017, voyles2018} and suggest the possibility of using
animals from these recovering populations to reestablish extirpated
populations. As with MYL frogs, the feasibility and long-term success of
such efforts will depend on the availability of robust donor populations
containing individuals that have acquired via evolution the adaptive
alleles necessary to allow frog survival and population growth in the
presence of Bd. Despite the hopeful example of successful
reestablishment of MYL frogs via evolutionary rescue and
reintroductions, the challenge of recovering hundreds of Bd-impacted
amphibians worldwide is a daunting prospect. Although we now have a
proven strategy to reestablish extirpated MYL frog populations, recovery
across their large historical range will require substantial resources
over many decades. The results of this study provide a hopeful starting
point for that endeavor, and other future efforts worldwide.

Evolutionary rescue may be increasingly critical for the survival of
wildlife populations in a rapidly changing world. The possibility - and
importance - of evolutionary rescue has been discussed for a wide
variety of taxa \citep[e.g.,][]{mills2018, oziolor2019}, and in the
specific context of disease-impacted systems \citep{searle2020}.
However, obtaining evidence in wild animal populations for all the
causal links in this process (i.e., selection for adaptive alleles
\(\rightarrow\) increase in host fitness in the presence of the pathogen
\(\rightarrow\) positive population growth and recovery) is difficult
and perhaps impossible. Nonetheless, we provide a compelling example
from the wild of how in the face of a novel stressor evolutionary rescue
can produce individuals that harbor adaptive alleles and that these
individuals can be used to reestablish extirpated imperiled populations.
We expect that such ``assisted evolutionary rescue'' will be an
essential tool in wildlife conservation in an era of global change.

@@

To visualize the genomic relationship of our populations we created a
PCA using the glPCA function in the adegenet R package
\citep{jombart2008}. Next, to detect regions of the genome under
selection between naive and recovering populations, we used a
multivariate linear mixed model to evaluate individual SNP variants and
a splined window analysis to evaluate larger genomic regions. For the
SNP analysis, we first used a stringent data filter to include only SNPs
with \textless{} 5\% missing data (missing for no more than 2
individuals), and then calculated the likelihood ratio statistic for the
resulting set of 148,307 high quality SNPs across 127 contigs using
GEMMA. GEMMA calculates and incorporates a relatedness matrix for input
samples, allowing us to account for relatedness and population structure
when calculating likelihood ratio statistics. We identified outlier SNPs
using a Bonferroni- corrected significance level of 0.01. Outliers are
SNPs showing different allele frequencies between naive versus
recovering populations. We visualized the results using a Manhattan plot
and qqplot. We developed a more liberal set of outlier SNPs using a
Bonferroni- corrected significance level of 0.05 and used this set
solely for the GO analysis (see below).

For each outlier SNP, we determined whether the variant was synonymous
(protein sequence the same for each variant) or non-synonymous (protein
sequence differs between variants), and where in the gene it was
located. To do this, we first extracted the reference genome sequence
surrounding the variant using bedtools ``getfasta'' \citep{quinlan2010}.
Next, we re-annotated each sequence using blast to get the predicted
gene location based on the closest annotated reference
\citep{altschul1997}. We then translated each variant to amino acids and
aligned this translation to that of the gene annotation to ensure proper
frame of reference using Geneious \citep{kearse2012}. After ensuring
proper translation, we characterized variants as within or outside the
coding sequence of the gene and as either synonymous or non-synonymous.

@@

Using the liberal set of variant outliers, we performed a statistical
overrepresentation test (Fisher's exact test) using Blast2GO
\citep{conesa2005}. First, we retrieved the Blast hits and mapped GO
terms for each gene in our targeted transcriptome. Next, we compared the
29 unique outlier genes from the liberal set of outlier variants to the
complete set of genes in our target transcriptome to determine if any GO
biological functions, molecular functions, or cellular processes were
overrepresented. We then repeated this process for the set of 33 genes
located in the overlapping \emph{F\textsubscript{ST}} and \(\pi_{diff}\)
splined windows.

@@

\caption{\label{fig-allelefrequencies}For each of the naive and
recovering MYL frog populations, pie charts showing allele frequencies
for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B)
TCF19, VARS, MAP1S, TYRP1, SRFBP1, and (C) RIN3. Charts are superimposed
on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon,
and Sequoia National Parks (from north to south) shown in dark gray, and
the range boundary for MYL frogs in this portion of the Sierra Nevada
shown in light gray. The inset map locates the study area in
California.}

@@

\caption{\label{fig-spline-manhattan}Results from the splined window
analysis showing outlier windows for the difference in nucleotide
diversity \(\pi_{diff}\) (A) and \emph{F\textsubscript{ST}} (B). Outlier
regions for either \(\pi_{diff}\) or \emph{F\textsubscript{ST}} are
shown in blue and outlier regions for both \(\pi_{diff}\) and
\emph{F\textsubscript{ST}} are shown in red. (C) Magnified Contig19 from
(A) showing two adjacent outlier windows for \(\pi_{diff}\) 12.9Mb
upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue
line). (D) Magnified Contig8 from from (B) showing the
\emph{F\textsubscript{ST}} outlier window that includes the outlier SNPs
TCF19 and VARS. This region of the genome contains 8 annotated genes
known to occur in the extended MHC Class I and II regions.}

@@

{\centering \includegraphics[width=0.8\textwidth]{figures/pop_viability_figures_for_manuscript.jpg}

>>>>

% Please add a significance statement to explain the relevance of your work
\significancestatement{Understanding how species persist despite accelerating global change is
critical for the conservation of biodiversity. One possibility is
evolutionary modification that restores positive growth to declining
populations (``evolutionary rescue''). Individuals from these rescued
populations could also be used to reestablish exinct populations. We
evaluated both possibilities in a once-common frog species driven to the
brink of extinction by a disease that has devastated global amphibian
biodiversity and for which few options exist to reverse its impacts. We
found that frogs have evolved increased resistance to infection,
facilitating population recovery. In addition, introduction of frogs
from these populations allowed reestablishment despite the ongoing
presence of disease. Such reintroductions could play an important role
in biodiversity conservation amidst unprecedented global change.}

@@

\begin{abstract}
Escalating anthropogenic change is devastating global biodiversity. To
avoid extinction, affected populations will often need to adapt to the
altered environment. This process of rapid evolutionary change that
increases the frequency of adaptive alleles and thereby restores
positive population growth is termed ``evolutionary rescue''. In
addition to allowing recovery of extant populations, individuals from
rescued populations could also be used to reestablish extirpated
populations. We evaluated this scenario using the once-common mountain
yellow-legged (MYL) frog, which is threatened with extinction by the
human-facilitated emergence of a lethal fungal pathogen
(\emph{Batrachochytrium dendrobatidis}; ``Bd''). Although most MYL frog
populations are extirpated following disease outbreaks, some persist and
eventually begin to recover. We show that MYL frogs have undergone
substantial evolutionary change following disease outbreaks, and that
these changes are associated with increased resistance/tolerance to Bd
infection in naturally recovering populations. Large-scale
reintroduction of frogs from these populations resulted in the
establishment of reproducing populations at most recipient sites despite
ongoing disease. In addition, results from viability modeling suggest
that established populations have a low probability of extinction over
50 years. Collectively, these results provide one of the first examples
of how evolutionary rescue and the reintroduction of resistant/tolerant
individuals can allow the landscape-scale recovery of disease-impacted
species. This example has potentially broad implications for the many
taxa worldwide that are threatened with extinction by anthropogenic
stressors.
\end{abstract}

@@

\dropcap{H}uman activities are intensifying threats to global biodiversity
\citep{ceballos2015}, with important implications for ecosystem
resilience and human welfare \citep{naeem2009}. To persist, affected
taxa will often need to adapt to their altered environment or face
decline and extinction. ``Evolutionary rescue'' describes rapid
evolutionary change that increases the frequency of adaptive alleles and
thereby restores positive population growth \citep{carlson2014}. In
addition to preventing extinction of declining populations, in theory,
this process can also result in the seeding of nearby empty habitats and
allow population reestablishment and recovery. Therefore, understanding
the mechanisms supporting population persistence, including evolutionary
rescue, is critical for the conservation of the ever-growing number of
threatened taxa.

One consequence of human alteration of the biosphere is an increase in
emerging infectious diseases \citep{jones2008, fisher2012}. Such
diseases pose a severe threat to wildlife populations
\citep{daszak2000}, and have caused dramatic declines and extinctions in
a wide range of taxa
\citep{hewson2014, samuel2015, scheele2019, cunningham2021}. Amid this
global loss of biodiverity, some populations previously decimated by
disease have subsequently recovered to varying degrees
\citep{newell2013, voyles2018, knapp2016}. Although we know relatively
little about the underlying mechanisms
\citep{brannelly2021, russell2020}, host evolution is one possibility
\citep{carlson2014, searle2020}. This process is known from a diversity
of wildlife populations
\citep{savage2016, epstein2016, gignoux-wolfsohn2021, holland2022}, and
often results in increased host resistance and/or tolerance (see
\citep{medzhitov2012} for definitions). However, linking these
evolutionary responses to population recoveries - as necessary for
evolutionary rescue - is difficult and few clear examples exist. In
addition to promoting recovery of disease-impacted populations, the
resistant/tolerant individuals in such populations could potentially be
used to reestablish populations that were extirpated following disease
emergence \citep[e.g.,][]{mendelson2019, brannelly2021}. To date, the
use of resistant/tolerant individuals for species recovery is rare
\citep[but see][]{joseph2018}, and studies that evaluate its potential
utility are critically needed.

The mountain yellow-legged (MYL) frog, composed of the sister species
\emph{Rana muscosa} and \emph{Rana sierrae} \citep{vredenburg2007}, is
emblematic of the unprecedented global declines of amphibians caused by
the recently-emerged amphibian chytrid fungus
\citep[\emph{Batrachochytrium dendrobatidis,}][]{scheele2019}. Once the
most common amphibian in the high elevation portion of California's
Sierra Nevada mountains \citep[USA,][]{grinnell1924}, this frog has
disappeared from more than 90\% of its historical range during the past
century \citep{vredenburg2007}. Due to the severity of its decline and
the increasing probability of extinction, both species are now listed as
``endangered'' under the U.S. Endangered Species Act. In the Sierra
Nevada, this decline was initiated by the introduction of non-native
trout into the extensive historically-fishless region
\citep{bradford1989, knapp2000}, and the arrival of \emph{B.
dendrobatidis} (Bd) in the 1960s and its subsequent spread
\citep{vredenburg2019} caused additional widespread population
extirpations \citep{vredenburg2010, rachowicz2006}. These Bd-caused
declines are fundamentally different from the fish-caused declines
because removal of fish populations is feasible and results in the rapid
recovery of frogs \citep{knapp2007, vredenburg2004}. In contrast, Bd
appears to persist in habitats even in the absence of amphibian hosts
\citep{walker2007}, and therefore represents a long-term alteration of
invaded ecosystems that amphibians will need to overcome to reestablish
populations.

Despite the catastrophic impact of Bd on MYL frogs, wherein most
Bd-naive populations are extirpated following Bd arrival
\citep{vredenburg2010}, some populations have persisted after epizootics
\citep{briggs2010} and are now recovering \citep{knapp2016}. Frogs in
these recovering populations show reduced susceptibility to Bd
infection, consistent with selection for more resistant/tolerant
genotypes \citep{knapp2016}. Furthermore, exploratory efforts to
reintroduce frogs collected from recovering populations to sites from
which frogs were previously extirpated indicate that population
establishment is possible despite ongoing Bd infection
\citep{joseph2018}.

In the current study, we combine results from genomic analyses, a
15-year landscape-scale frog reintroduction effort, and detailed
population models to provide an unprecedented example of how evolution
in response to disease allows the successful reestablishment of
extirpated MYL frog populations. Specifically, we used exome capture
methods to show that MYL frogs have undergone substantial evolutionary
change associated with Bd exposure and that may provide a genetic basis
for the observed increased resistance/tolerance to Bd infection observed
in naturally recovering populations \citep{knapp2016}. We then describe
the reintroduction of these resistant/tolerant frogs to 24 sites and the
dynamics and drivers of population establishment using results from
long-term capture-mark-recapture (CMR) surveys. These data are also used
to parameterize a population model that extends our inferences of
population establishment well beyond the temporal extent of the CMR
data. Collectively, these results provide one of the first examples of
how evolutionary rescue and the reintroduction of resistant/tolerant
animals can allow the large-scale recovery of disease-impacted species.

@@

Even when incorporating demographic stochasticity and environmental
stochasticity in the probability of juvenile recruitment to the adult
class \(\omega\) (the transition that we expect to be the most subject
to environmental variability in the presence of Bd), populations with
high adult survival are likely to persist over a 50 year time horizon.
Our model predicted that, following a single introduction of 40 adult
individuals into a population and using the baseline value for
\(\omega\) of 0.3, the six populations with the highest adult survival
probabilities had 50-year extinction probabilities of 0 to 0.32;
Figure~\ref{fig-viability} B). This indicates strong potential for
long-term persistence in the presence of Bd and environmental
variability in recruitment (Figure~\ref{fig-viability} B). When we
decreased \(\omega\) from 0.3 to a conservative estimate of 0.15,
50-year extinction probability remained less than 0.5 for three
populations and was 0.79 to 0.94 for three additional populations
(Figure~\ref{fig-viability} B). In contrast, for the six populations
where \(\sigma_{A_R} < 0.5\), extinction probability over 50 years was
always predicted to be 1. To test the validity of our model predictions,
we demonstrated that our stochastic model could describe the general
recovery trajectory of our translocated population with the longest
survey history (Figure~\ref{fig-viability} C; population 70550, surveyed
for 16 years).

@@

Disease-induced population declines are decimating global biodiversity
\citep{daszak2000}, but strategies to recover affected species are
generally lacking. Here, we used populations of resistant/tolerant
individuals to reestablish extirpated populations, despite ongoing
disease. Although previously proposed \citep{mendelson2019}, this
strategy has rarely been successfully implemented (see
\citep{joseph2018} for a small-scale example). In particular, although
numerous studies have documented signatures of selection in hosts
following disease-induced declines
\citep[e.g.,][]{epstein2016, gignoux-wolfsohn2021}, we are unaware of
any that used individuals bearing the resulting adaptive alleles to
reestablish extirpated populations. The results of our study provide
evidence that (i) MYL frogs have experienced strong selection following
exposure to a recently-emerged lethal fungal disease and that some of
the genomic regions under selection are associated with genes related to
disease resistance/tolerance, and (ii) reintroduction of individuals
from these resistant/tolerant populations allows the successful
reestablishment of viable frog populations despite ongoing Bd infection.

@@

Results from our genomic analyses strongly suggest that natural
selection for adaptive alleles and resulting increased
resistance/tolerance is at least partially responsible for the reduced
susceptibility of frogs in recovering populations. We identified
specific alleles and genomic regions showing signatures of selection
between adjacent Bd-naive and recovering MYL frog populations,
consistent with strong selection following Bd exposure. Importantly,
some regions under selection are associated with cellular and
immunological mechanisms known to contribute to disease resistance,
including in amphibians \citep{zamudio2020}. For example, the MHC plays
an important role in immunity. In our study, we identified a region that
shows evidence of selection in recovering populations and contains eight
genes associated with either the MHC Class I or Class III regions. These
results corroborate numerous previous studies linking MHC genes to
amphibian resistance/tolerance against Bd
\citep[e.g.,][]{savage2011, bataille2015}. Similarly, the region with
the strongest indication of directional selection (as measured by
\(\pi_{diff}\)) contains the gene GVINP1. Several previous studies have
found this gene to be differentially expressed during Bd infection
\citep[e.g.,][]{grogan2018, ellison2014}. This gene is also strongly
linked to disease in salmon, explaining a notable 20\% of the resistance
phenotype \citep{robledo2020, robledo2018}. Finally, the RIN3 gene
showed a consistent pattern of allele frequency differences across all
nine of the frog populations sampled in this study, indicating
consistent selection in populations distributed across a wide geographic
area. This gene is associated with inflammation response and in
\emph{Xenopus} is highly expressed in wound stumps following amputation
of tadpole tails \citep{fukazawa2009}. In the SI, we provide additional
details on these and other genes identified in our study as under
selection.

@@

Despite the resistance/tolerance of adult MYL frogs in recovering
populations, individual and population-level impacts of Bd are still
evident. In an earlier study of 2 of our 12 translocated populations, Bd
infection and load had detectable effects on the survival of adults and
may have influenced population establishment (sites referred to as
``Alpine'' and ``Subalpine'' in \citep{joseph2018} are identified as
``70550'' and ``70505'' in the current study). Applying similar analyses
to all 12 of our translocated populations would likely provide a broader
perspective of the ongoing effect of Bd. In addition to these important
but relatively subtle effects on adults, MYL frogs immediately following
metamorphosis (``metamorphs'') are highly susceptible to Bd infection
\citep{ellison2018} and experience high mortality. This high
susceptibility of metamorphs is documented for numerous species of
anurans, and may result from the poorly developed immune system
characteristic of this early life stage \citep{humphries2022}. In
naturally recovering and translocated MYL frog populations, we suggest
that the high mortality of metamorphs is an important limitation on
subsequent recruitment of new adults. Therefore, although adult MYL
frogs appear relatively resistant/tolerant, Bd infection continues to
have important limiting effects on recovering populations \citep[see
also][]{hollanders2022}.

The recent emergence of Bd worldwide has contributed to the decline of
hundreds of amphibian species, some of which are now extinct in the wild
\citep{scheele2019}. This unprecedented impact on global amphibian
biodiversity is compounded by the lack of any effective and broadly
applicable strategies to reverse these impacts
\citep{garner2016, knapp2022}. Importantly, in addition to the natural
recovery documented for MYL frogs \citep{knapp2016}, other species are
also showing evidence of post-epizootic recovery in the presence of Bd
\citep{scheele2017, voyles2018} and suggest the possibility of also
using animals from these recovering populations to reestablish
extirpated populations. As with MYL frogs, the feasibility and long-term
success of such efforts will depend on the availability of robust donor
populations containing individuals that have acquired via evolution the
adaptive alleles necessary to allow frog survival and population growth
in the presence of Bd. Despite the hopeful example of successful
reestablishment of MYL frogs via evolutionary rescue and
reintroductions, the challenge of recovering hundreds of Bd-impacted
amphibian species worldwide is a daunting prospect. Although we now have
a proven strategy to reestablish extirpated MYL frog populations,
recovery across their large historical range will require substantial
resources over many decades. The results of this study provide a hopeful
starting point for that endeavor, and other future efforts worldwide.

Evolutionary rescue may be increasingly critical for the survival of
wildlife populations in a rapidly changing world. The importance of
evolutionary rescue is known for a wide variety of taxa
\citep[e.g.,][]{carlson2014, mills2018, oziolor2019}, including in the
specific context of disease-impacted systems \citep{searle2020}.
However, obtaining evidence in wild animal populations for all the
causal links in this process (i.e., selection for adaptive alleles
\(\rightarrow\) increase in host fitness in the presence of the stressor
\(\rightarrow\) positive population growth and recovery) is difficult in
most systems. Nonetheless, we provide a compelling example from the wild
not only that evolutionary rescue can produce individuals that harbor
adaptive alleles and allow population recovery, but importantly that
these individuals can be used in translocations to reestablish
extirpated populations. We expect that such ``assisted evolutionary
rescue'' will be an essential tool in wildlife conservation in an era of
accelerating global change.

@@

To visualize the genomic relationship of our populations we conducted a
principal components analysis (PCA) using the glPCA function in the
adegenet R package \citep{jombart2008}. To detect regions of the genome
under selection between naive and recovering populations, we used two
approches: (1) a multivariate linear mixed model to evaluate individual
variants (SNPs and INDELs), and (2) a splined window analysis to
evaluate larger genomic regions. For the variant analysis, we first used
a stringent data filter to include only variants with \textless{} 5\%
missing data (missing for no more than 2 individuals), and then
calculated the likelihood ratio statistic for the resulting set of
148,307 high quality variants across 127 contigs using GEMMA. GEMMA
calculates and incorporates a relatedness matrix for input samples,
allowing us to account for relatedness and population structure when
calculating likelihood ratio statistics. We identified ``outlier''
variants using a Bonferroni- corrected significance level of 0.01.
Outliers are variants showing different allele frequencies between naive
versus recovering populations. We visualized the results using a
Manhattan plot and qqplot. We developed a more liberal set of outlier
variants using a Bonferroni- corrected significance level of 0.05 and
used this set solely for the gene ontology (GO) analysis (see below).

For each outlier variant, we determined whether the variant was
synonymous (protein sequence the same for each variant) or
non-synonymous (protein sequence differs between variants), and where in
the gene it was located. To do this, we first extracted the reference
genome sequence surrounding the variant using the bedtools ``getfasta''
function \citep{quinlan2010}. Next, we re-annotated each sequence using
blast to get the predicted gene location based on the closest annotated
reference \citep{altschul1997}. We then translated each variant to amino
acids and aligned this translation to that of the gene annotation to
ensure proper frame of reference using Geneious \citep{kearse2012}.
After ensuring proper translation, we characterized variants as within
or outside the coding sequence of the gene and as either synonymous or
non-synonymous.

@@

Using the liberal set of variant outliers, we performed a statistical
overrepresentation test (Fisher's exact test) using Blast2GO
\citep{conesa2005}. First, we retrieved the Blast hits and mapped GO
terms for each gene in our targeted transcriptome. Next, we compared the
29 unique outlier genes from the liberal set of outlier variants to the
complete set of genes in our target transcriptome to determine if any GO
biological functions, molecular functions, or cellular processes were
overrepresented. We then repeated this process for the set of 33 genes
located in the overlapping \emph{F\textsubscript{ST}} and \(\pi_{diff}\)
splined windows. Code used for genomic analyses and to figures is
available from the following GitHub repository:
\url{https://github.com/allie128/mylf-selection}.

@@

\caption{\label{fig-allelefrequencies}Evidence for selection in
recovering MYL frog populations at the landscape scale. For each of the
naive and recovering MYL frog populations, pie charts show allele
frequencies for the 11 outlier SNPs from 7 distinct genes: (A)
LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, SRFBP1, and (C) RIN3.
Charts are superimposed on a map of the Sierra Nevada study area, with
Yosemite, Kings Canyon, and Sequoia National Parks (from north to south)
shown in dark gray, and the range boundary for MYL frogs in this portion
of the Sierra Nevada shown in light gray. The inset map locates the
study area in California.}

@@

\caption{\label{fig-spline-manhattan}Evidence for selection on genomic
regions in recovering MYL frog populations. Results are from the splined
window analysis and show outlier windows for the difference in (A)
nucleotide diversity \(\pi_{diff}\) and (B) \emph{F\textsubscript{ST}}.
Outlier regions for either \(\pi_{diff}\) or \emph{F\textsubscript{ST}}
are shown in blue and outlier regions for both \(\pi_{diff}\) and
\emph{F\textsubscript{ST}} are shown in red. (C) Magnified Contig19 from
(A) showing two adjacent outlier windows for \(\pi_{diff}\) 12.9Mb
upstream of the RIN3 outlier SNP (indicated with a dashed vertical blue
line). (D) Magnified Contig8 from from (B) showing the
\emph{F\textsubscript{ST}} outlier window that includes the outlier SNPs
TCF19 and VARS. This region of the genome contains 8 annotated genes
known to occur in the extended MHC Class I and II regions.}

@@

{\centering \includegraphics[width=\textwidth]{figures/pop_viability_figures_for_manuscript.jpg}
